Quotient maps and configuration spaces of hard disks

### 2022-02-09 manuscript/main.tex
responded comments 5

As stated previously, the configuration space in Eq.\ \ref{eq:configuration_space2} contains redundant information. Specifically, every configuration is equivalent to multiple other configurations related by the symmetry operations discussed by Ritchey \cite{ritcheyphd}. As mentioned in the Introduction, quotienting the configuration space by the translation group is so widespread that this operation is often not explicitly mentioned. The motivation to do so is that the resulting configuration space is much smaller than the base configuration space. That said, the quotient maps are such that it is often not clear how to explicitly parameterize the quotient spaces, though this would certainly facilitate the construction of an explicit triangulation. This section describes our procedure to do so.

@@

All other quotient spaces are initially embedded in an infinite-dimensional descriptor space, and a numerical approach is used to estimate the minimum number of descriptors necessary to maintain the embedding. \red{The idea is analogous to truncating a Fourier series representation of a function instead of keeping the infinitely many coefficients.} Given a configuration space of $n$ disks, the distribution $f$ is defined as a sum of Dirac-delta distributions $\delta(\vec{a}_j)$ located at the disk centers $\vec{a}_j$ in the $a_1a_2$-coordinate system in Fig.\ \ref{fig:figure2}, or
\begin{equation}
f(\vec{a}) = \sum_{j = 1}^{n} \delta(\vec{a}_j) = \sum_{\vec{k}} {c_{\vec{k}} \exp{(2\pi \rmi \vec{k} \cdot \vec{a})}}
\label{eq:f}
\end{equation}
where $c_{\vec{k}}$ are the complex coefficients of the reciprocal lattice expansion and $\vec{k} = [p, q]$ for integers $p$ and $q$. The infinite set of $c_{\vec{k}}$ can be calculated using the orthogonality of the complex exponentials as
\begin{equation}
c_{\vec{k}} = \sum_{j=1}^{n} \exp{(-2\pi \rmi \vec{k} \cdot \vec{a}_j)} .
\label{eq:c_k}
\end{equation}
$c_{\vec{k}}$ respects the periodicity of the lattice and is invariant to permutations of the disk labels due to the commutative property of the summation in Eq.\ \ref{eq:f}. It can be shown that translating a configuration (by adding an offset to the $\vec{a}_j$) only changes the phase of the coefficients. This means that the moduli of the coefficients, or
\begin{equation}
z_{\vec{k}} = \sqrt{c^\ast_{\vec{k}} c_{\vec{k}}}
\label{eq:z_k}
\end{equation}
where $^\ast$ denotes the complex conjugate, are a set of real-valued descriptors that are invariant to disk label permutations and rigid translations. Observe that the descriptors $z_{\vec{k}}$ also respect inversion symmetry. Numerical experiments indicate that the rank of the Jacobian of the map from the $\vec{a}_j$ to the $z_{\vec{k}}$ is generically $2(n - 1)$, suggesting that some number of these descriptors could be sufficient to construct an embedding of $\Lambda / \mathcal{S}_2$.

@@

\begin{figure*}
	\centering
	\includegraphics[width=1.0\textwidth]{figure9.eps}
	\caption{The evolution of the translation invariant configuration space $\Gamma(2, \rho) / \mathcal{S}_1$ for the square torus (top) and for the hexagonal torus (bottom) with $\rho=\{0.28, 0.26, 0.25, 0.21, 0.17, 0.12\}$. The locations of the critical points in this space are indicated by arrows.}
	\label{fig:figure9}
\end{figure*}

Figure \ref{fig:figure9} shows the translation-invariant configuration space $\Gamma(2, \rho) / \mathcal{S}_1$ of two disks as a function of $\rho$ for the square torus (top) and hexagonal torus (bottom) as obtained from the $\alpha$-complex of 10 000 points. Note that the square torus is constructed by extending the square to a rectangle and identifying opposite edges, but this does not affect the topological properties of the space. When $\rho > 0.25$, the space $\Gamma(2, \rho) / \mathcal{S}_1$ is comprised of a single $0$-handle whereas that of the hexagonal torus is comprised of two $0$-handles. This difference should be expected on the basis of Fig.\ \ref{fig:figure8} since the two index-$0$ critical points of the hexagonal torus are not related by a rigid translation. When $\rho = 0.25$, two and three $1$-handles are connected for the square and the hexagonal toruses, respectively. Observe that the $1$-handles provide connections between previously distant regions of the space. For $\rho < 0.25$, the space continues to grow and eventually closes in the $\rho \rightarrow 0$ limit. That is, the configuration with $\rho = 0$ acts like an index-$2$ critical point, even though it is not strictly within the space.

@@

The descriptors $z_{\vec{k}}$ defined in Sec.\ \ref{sec:descriptors} are by construction invariant to rigid translations, inversions about the origin, and permutations of disk labels. One way to construct the quotient space $\Lambda / \mathcal{S}_2$ is then to use the $z_{\vec{k}}$ as coordinates for the descriptor space. Figure \ref{fig:figure6} suggests that four of these are sufficient for a proper embedding of $\Lambda / \mathcal{S}_2$. The ISOMAP algorithm is used to reduce the dimension further by one, allowing visualization of the quotient space, but requires a distance function to do so. The top rows of Fig.\ \ref{fig:figure10} and Fig.\ \ref{fig:figure11} use the Euclidean distance in the descriptor space, whereas the bottom rows use the distance defined in Eq.\ \ref{eq:unlabeled_distance}. This allows two versions of $\Gamma(2, \rho) / \mathcal{S}_2$ to be constructed for both the square and hexagonal toruses; it is significant that the two versions are topologically identical, though the one using Eq.\ \ref{eq:unlabeled_distance} better preserves the expected quotient space symmetries; analogous to the truncation of a Fourier series, the use of a distance based on a finite number of descriptors likely introduces distortions. Regardless, $\Gamma(2, \rho) / \mathcal{S}_2$ starts with the index-$0$ critical points and grows without topological change until $\rho = 0.25$ when the index-$1$ critical points appear. Unlike for $\Gamma(2, \rho) / \mathcal{S}_1$, these critical points don't appear as handles, but as singular points.

@@

\begin{figure*}
	\centering
	\includegraphics[width=1.0\textwidth]{figure10.eps}
	\caption{The evolution of the translation, permutation and inversion invariant configuration space $\Gamma(2, \rho) / \mathcal{S}_2$ for the square torus constructed with the standard Euclidean distance (top) and the distance in Eq.\ \ref{eq:unlabeled_distance} (bottom) with $\rho = \{0.28, 0.26, 0.25, 0.21, 0.17, 0.12\}$. The locations of the critical points are indicated by arrows.}
	\label{fig:figure10}
\end{figure*}

\begin{figure*}
	\centering
	\includegraphics[width=1.0\textwidth]{figure11.eps}
	\caption{The evolution of the translation, permutation and inversion invariant configuration space $\Gamma(2, \rho) / \mathcal{S}_2$ for the hexagonal torus constructed with the standard Euclidean distance (top) and the distance in Eq.\ \ref{eq:unlabeled_distance} (bottom) with $\rho = \{0.28, 0.26, 0.25, 0.21, 0.17, 0.12\}$. The locations of the critical points are indicated by arrows.}
	\label{fig:figure11}
\end{figure*}

@@

The descriptors $\hat{z}_{\vec{k}}$ defined in Sec.\ \ref{sec:descriptors} are additionally invariant to the symmetries of the plane tiling, and are used as coordinates for the embedding of the quotient space $\Lambda / \mathcal{S}_3$. As before, dimensionality reduction is performed with the ISOMAP algorithm. The top rows of Fig.\ \ref{fig:figure12} and Fig.\ \ref{fig:figure13} use the Euclidean distance in the descriptor space, whereas the bottom rows use the distance defined in Eq.\ \ref{eq:unlabeled_distance}. The two versions of $\Gamma(2, \rho) / \mathcal{S}_3$ are topologically identical as before. That said, the one using Eq.\ \ref{eq:unlabeled_distance} better preserves the expected quotient space symmetries, with the geometric distortions introduced by using the Euclidean distance in the descriptor space much more pronounced than those in Fig.\ \ref{fig:figure10} and Fig.\ \ref{fig:figure11}. Specifically, the version of $\Gamma(2, \rho) / \mathcal{S}_3$ constructed with the Euclidean distance incorrectly collapses the region for small $\rho$ to a $1$-manifold. Further examination suggests that the quotient spaces constructed with Eq.\ \ref{eq:unlabeled_distance} are the smallest symmetric regions of their corresponding domains; the bottom row of Fig.\ \ref{fig:figure12} is $1 / 8$ of the square torus, whereas that of Fig.\ \ref{fig:figure13} is $1 / 12$ of the hexagonal torus. The corresponding fundamental cells can be obtained by reflecting the quotient spaces along an edge passing through the $\rho = 0$ point and applying the appropriate rotations.

@@

\begin{figure*}
	\centering
	\includegraphics[width=1.0\textwidth]{figure12.eps}
	\caption{The evolution of the translation, permutation, inversion and lattice symmetry invariant configuration space $\Gamma(2, \rho) / \mathcal{S}_3$ for the square torus constructed with the standard Euclidean distance (top) and with the distance in Eq.\ \ref{eq:unlabeled_distance} (bottom) with $\rho=\{0.28, 0.26, 0.25, 0.21, 0.17, 0.12\}$. The locations of the critical points are indicated by arrows.}
	\label{fig:figure12}
\end{figure*}

\begin{figure*}
	\centering
	\includegraphics[width=1.0\textwidth]{figure13.eps}
	\caption{The evolution of the translation, permutation, inversion and lattice symmetry invariant configuration space $\Gamma(2, \rho) / \mathcal{S}_3$ for the hexagonal torus constructed with the standard Euclidean distance (top) and with the distance in Eq.\ \ref{eq:unlabeled_distance} (bottom) with $\rho=\{0.28, 0.26, 0.25, 0.21, 0.17, 0.12\}$. The locations of the critical points are indicated by arrows.}
	\label{fig:figure13}
\end{figure*}

>>>>

As stated previously, the configuration space in Eq.\ \ref{eq:configuration_space2} contains redundant information. Specifically, every configuration is equivalent to multiple other configurations related by the symmetry operations discussed by Ritchey \cite{ritcheyphd}. As mentioned in the Introduction, quotienting the base configuration space by the translation group is so widespread that this operation is often not explicitly mentioned. The motivation to do so is that the resulting quotient space is much smaller than the base configuration space. That said, the quotient maps are such that it is often not clear how to explicitly parameterize the quotient spaces, though this would certainly facilitate the construction of an explicit triangulation. This section describes our procedure to do so.

@@

All other quotient spaces are initially embedded in an infinite-dimensional descriptor space, and a numerical approach is used to estimate the minimum number of descriptors necessary to maintain the embedding. \red{This idea is analogous to representing a function in the Fourier space, but using a finite subset of infinitely-many coefficients.} Given a configuration space of $n$ disks, the distribution $f$ is defined as a sum of Dirac-delta distributions $\delta(\vec{a}_j)$ located at the disk centers $\vec{a}_j$ in the $a_1a_2$-coordinate system in Fig.\ \ref{fig:figure2}, or
\begin{equation}
f(\vec{a}) = \sum_{j = 1}^{n} \delta(\vec{a}_j) = \sum_{\vec{k}} {c_{\vec{k}} \exp{(2\pi \rmi \vec{k} \cdot \vec{a})}}
\label{eq:f}
\end{equation}
where $c_{\vec{k}}$ are the complex coefficients of the reciprocal lattice expansion and $\vec{k} = [p, q]$ for integers $p$ and $q$. The infinite set of $c_{\vec{k}}$ can be calculated using the orthogonality of the complex exponentials as
\begin{equation}
c_{\vec{k}} = \sum_{j=1}^{n} \exp{(-2\pi \rmi \vec{k} \cdot \vec{a}_j)} .
\label{eq:c_k}
\end{equation}
$c_{\vec{k}}$ respects the periodicity of the lattice and is invariant to permutations of the disk labels due to the commutative property of the summation in Eq.\ \ref{eq:f}. It can be shown that translating a configuration (by adding an offset to the $\vec{a}_j$) only changes the phase of the coefficients. This means that the moduli of the coefficients, or
\begin{equation}
z_{\vec{k}} = \sqrt{c^\ast_{\vec{k}} c_{\vec{k}}}
\label{eq:z_k}
\end{equation}
where $^\ast$ denotes the complex conjugate, are a set of real-valued descriptors that are invariant to disk label permutations and rigid translations. Observe that the descriptors $z_{\vec{k}}$ also respect inversion symmetry. Numerical experiments indicate that the rank of the Jacobian of the map from the $\vec{a}_j$ to the $z_{\vec{k}}$ is generically $2(n - 1)$, suggesting that some number of these descriptors could be sufficient to construct an embedding of $\Lambda / \mathcal{S}_2$.

@@

\begin{figure*}
	\centering
	\includegraphics[width=1.0\textwidth]{figure9.eps}
	\caption{The evolution of the translation invariant configuration space, \red{or the quotient space $\Gamma(2, \rho) / \mathcal{S}_1$,} for the square torus (top) and for the hexagonal torus (bottom) with $\rho=\{0.28, 0.26, 0.25, 0.21, 0.17, 0.12\}$. The locations of the critical points in this space are indicated by arrows.}
	\label{fig:figure9}
\end{figure*}

Figure \ref{fig:figure9} shows the translation-invariant configuration space \red{(or quotient space $\Gamma(2, \rho) / \mathcal{S}_1$)} of two disks as a function of $\rho$ for the square torus (top) and hexagonal torus (bottom) as obtained from the $\alpha$-complex of 10 000 points. Note that the square torus is constructed by extending the square to a rectangle and identifying opposite edges, but this does not affect the topological properties of the space. When $\rho > 0.25$, the space $\Gamma(2, \rho) / \mathcal{S}_1$ is comprised of a single $0$-handle whereas that of the hexagonal torus is comprised of two $0$-handles. This difference should be expected on the basis of Fig.\ \ref{fig:figure8} since the two index-$0$ critical points of the hexagonal torus are not related by a rigid translation. When $\rho = 0.25$, two and three $1$-handles are connected for the square and the hexagonal toruses, respectively. Observe that the $1$-handles provide connections between previously distant regions of the space. For $\rho < 0.25$, the space continues to grow and eventually closes in the $\rho \rightarrow 0$ limit. That is, the configuration with $\rho = 0$ acts like an index-$2$ critical point, even though it is not strictly within the space.

@@

The descriptors $z_{\vec{k}}$ defined in Sec.\ \ref{sec:descriptors} are by construction invariant to rigid translations, inversions about the origin, and permutations of disk labels. One way to construct the quotient space $\Lambda / \mathcal{S}_2$ is then to use the $z_{\vec{k}}$ as coordinates. Figure \ref{fig:figure6} suggests that four of these are sufficient for a proper embedding of $\Lambda / \mathcal{S}_2$. The ISOMAP algorithm is used to reduce the dimension further by one, allowing visualization of the quotient space, but requires a distance function to do so. The top rows of Fig.\ \ref{fig:figure10} and Fig.\ \ref{fig:figure11} use the Euclidean distance \red{using the corresponding descriptors}, whereas the bottom rows use the distance defined in Eq.\ \ref{eq:unlabeled_distance}. This allows two versions of \red{the quotient space} $\Gamma(2, \rho) / \mathcal{S}_2$ to be constructed for both the square and hexagonal toruses; it is significant that the two versions are topologically identical, though the one using Eq.\ \ref{eq:unlabeled_distance} better preserves the expected quotient space symmetries; analogous to the truncation of a Fourier series, the use of a distance based on a finite number of descriptors likely introduces distortions. Regardless, $\Gamma(2, \rho) / \mathcal{S}_2$ starts with the index-$0$ critical points and grows without topological change until $\rho = 0.25$ when the index-$1$ critical points appear. Unlike for $\Gamma(2, \rho) / \mathcal{S}_1$, these critical points don't appear as handles, but as singular points.

@@

\begin{figure*}
	\centering
	\includegraphics[width=1.0\textwidth]{figure10.eps}
	\caption{The evolution of the translation, permutation and inversion invariant configuration space, \red{or the quotient space $\Gamma(2, \rho) / \mathcal{S}_2$,} for the square torus constructed with the standard Euclidean distance (top) and the distance in Eq.\ \ref{eq:unlabeled_distance} (bottom) with $\rho = \{0.28, 0.26, 0.25, 0.21, 0.17, 0.12\}$. The locations of the critical points are indicated by arrows.}
	\label{fig:figure10}
\end{figure*}

\begin{figure*}
	\centering
	\includegraphics[width=1.0\textwidth]{figure11.eps}
	\caption{The evolution of the translation, permutation and inversion invariant configuration space, \red{or the quotient space $\Gamma(2, \rho) / \mathcal{S}_2$,} for the hexagonal torus constructed with the standard Euclidean distance (top) and the distance in Eq.\ \ref{eq:unlabeled_distance} (bottom) with $\rho = \{0.28, 0.26, 0.25, 0.21, 0.17, 0.12\}$. The locations of the critical points are indicated by arrows.}
	\label{fig:figure11}
\end{figure*}

@@

The descriptors $\hat{z}_{\vec{k}}$ defined in Sec.\ \ref{sec:descriptors} are additionally invariant to the symmetries of the plane tiling, and are used as coordinates for the embedding of the quotient space $\Lambda / \mathcal{S}_3$. As before, dimensionality reduction is performed with the ISOMAP algorithm. The top rows of Fig.\ \ref{fig:figure12} and Fig.\ \ref{fig:figure13} use the Euclidean distance \red{among the descriptors}, whereas the bottom rows use the distance defined in Eq.\ \ref{eq:unlabeled_distance}. The two versions of $\Gamma(2, \rho) / \mathcal{S}_3$ are topologically identical as before. That said, the one using Eq.\ \ref{eq:unlabeled_distance} better preserves the expected quotient space symmetries, with the geometric distortions introduced by using the Euclidean distance in the descriptor space much more pronounced than those in Fig.\ \ref{fig:figure10} and Fig.\ \ref{fig:figure11}. Specifically, the version of $\Gamma(2, \rho) / \mathcal{S}_3$ constructed with the Euclidean distance incorrectly collapses the region for small $\rho$ to a $1$-manifold. Further examination suggests that the quotient spaces constructed with Eq.\ \ref{eq:unlabeled_distance} are the smallest symmetric regions of their corresponding domains; the bottom row of Fig.\ \ref{fig:figure12} is $1 / 8$ of the square torus, whereas that of Fig.\ \ref{fig:figure13} is $1 / 12$ of the hexagonal torus. The corresponding fundamental cells can be obtained by reflecting the quotient spaces along an edge passing through the $\rho = 0$ point and applying the appropriate rotations.

@@

\begin{figure*}
	\centering
	\includegraphics[width=1.0\textwidth]{figure12.eps}
	\caption{The evolution of the translation, permutation, inversion and lattice symmetry invariant configuration space, \red{or the quotient space $\Gamma(2, \rho) / \mathcal{S}_3$,} for the square torus constructed with the standard Euclidean distance (top) and with the distance in Eq.\ \ref{eq:unlabeled_distance} (bottom) with $\rho=\{0.28, 0.26, 0.25, 0.21, 0.17, 0.12\}$. The locations of the critical points are indicated by arrows.}
	\label{fig:figure12}
\end{figure*}

\begin{figure*}
	\centering
	\includegraphics[width=1.0\textwidth]{figure13.eps}
	\caption{The evolution of the translation, permutation, inversion and lattice symmetry invariant configuration space, \red{or the quotient space $\Gamma(2, \rho) / \mathcal{S}_3$,} for the hexagonal torus constructed with the standard Euclidean distance (top) and with the distance in Eq.\ \ref{eq:unlabeled_distance} (bottom) with $\rho=\{0.28, 0.26, 0.25, 0.21, 0.17, 0.12\}$. The locations of the critical points are indicated by arrows.}
	\label{fig:figure13}
\end{figure*}
